# Overview
Monitoring the distribution and change in land cover types can help us understand the impacts of phenomena like climate change, natural disasters, deforestation, and urbanization. Determining land cover types over large areas is a major application of remote sensing because we are able to distinguish different materials based on their spectral reflectance. 

Classifying remotely sensed imagery into landcover classes enables us to understand the distribution and change in landcover types over large areas. There are many approaches for performing landcover classification -- *supervised* approaches use training data labeled by the user, whereas *unsupervised* approaches use algorithms to create groups which are identified by the user afterward.\

credit: this lab is based on a materials developed by Chris Kibler.

# Task
In this lab, we are using a form of supervised classification, a *decision tree classifier*. [Decision trees](https://medium.com/@ml.at.berkeley/machine-learning-crash-course-part-5-decision-trees-and-ensemble-models-dcc5a36af8cd) classify pixels using a series of conditions based on values in spectral bands. These conditions (or decisions) are developed based on training data. In this lab we will create a land cover classification for southern Santa Barbara County based on multi-spectral imagery and data on the location of 4 land cover types:

-   green vegetation\
-   dry grass or soil\
-   urban\
-   water\

## Summary

-   load and process Landsat scene\
-   crop and mask Landsat data to study area\
-   extract spectral data at training sites\
-   train and apply decision tree classifier\
-   plot results

## Data

**Landsat 5 Thematic Mapper**\

-   [Landsat 5](https://www.usgs.gov/landsat-missions/landsat-5)
-   1 scene from September 25, 2007\
-   bands: 1, 2, 3, 4, 5, 7
-   Collection 2 surface reflectance product\

**Study area and training data**

-   polygon representing southern Santa Barbara county
-   polygons representing training sites\
    - type: character string with land cover type\

# Workflow
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
## Process data
#### Load packages and set working directory
We'll be working with vector and raster data, so will need both `sf` and `terra`. To train our classification algorithm and plot the results, we'll use the `rpart` and `rpart.plot` packages.
Set your working directory to the folder that holds the data for this lab.\

**Note:** my filepaths may look different than yours!
```{r include=TRUE, message=FALSE, warning=FALSE}
library(sf)
library(terra)
library(here)
library(dplyr)
library(rpart)
library(rpart.plot)
library(tmap)

rm(list = ls())

# here::i_am("labs/week9.Rmd")
setwd(here())
```

#### Load Landsat data
Let's create a raster stack based on the 6 bands we will be working with. Each file name ends with the band number (e.g. `B1.tif`). Notice that we are missing a file for band 6. Band 6 corresponds to thermal data, which we will not be working with for this lab. To create a raster stack, we will create a list of the files that we would like to work with and read them all in at once using the `rast` function. We'll then update the names of the layers to match the spectral bands and plot a true color image to see what we're working with.

```{r include=TRUE}
#list files for each band, including file path
filelist <- list.files("./data (1)/landsat-data", full.names = TRUE)
```

#### Load study area
We want to contstrain our analysis to the southern portion of the county where we have training data, so we'll read in a file that defines the area we would like to study.

```{r include=TRUE}
landsat_20070925 <- rast(filelist)

names(landsat_20070925) <- c("blue", "green", "red", "NIR", "SWIR1", "SWIR2")

#plot true color image
plotRGB(landsat_20070925, r = 3, g = 2, b = 1, stretch = "lin")
```

#### Crop and mask Landsat data to study area
Now, we can crop and mask the Landsat data to our study area. This reduces the amount of data we'll be working with and therefore saves computational time. We can also remove any objects we're no longer working with to save space. 
```{r include=TRUE}
# read in shapefile for southern portion of SB county

SB_county_south <- st_read("./data (1)/SB_county_south.shp")

SB_county_south <- st_transform(SB_county_south, crs = crs(landsat_20070925))

landsat_cropped <- crop(landsat_20070925, SB_county_south) #crop landsat to extend of SB_county shape file
plot(landsat_cropped)

landsat_masked <- mask(landsat_cropped, SB_county_south) #keeps values for only where SB_county is 
plotRGB(landsat_masked, r = 3, g = 2, b = 1, stretch = "lin")
```

#### Convert Landsat values to reflectance
Now we need to convert the values in our raster stack to correspond to reflectance values. To do so, we need to remove erroneous values and apply any [scaling factors](https://www.usgs.gov/faqs/how-do-i-use-scale-factor-landsat-level-2-science-products#:~:text=Landsat%20Collection%202%20surface%20temperature,the%20scale%20factor%20is%20applied.) to convert to reflectance.\
g
In this case, we are working with [Landsat Collection 2](https://www.usgs.gov/landsat-missions/landsat-collection-2). The valid range of pixel values for this collection 7,273-43,636, with a multiplicative scale factor of 0.0000275 and an additive scale factor of -0.2. So we reclassify any erroneous values as `NA` and update the values for each pixel based on the scaling factors. Now the pixel values should range from 0-100%.

```{r include=TRUE}
#correct range 7273 - 43636, replace any values that fall outside of that with NA
#^to get these numbers, google landsat5 scale factor or landsat5 specifications or something like that (linked above)

#define a matrix to replace bad values with NA
rcl_0 <- matrix(c(-Inf, 7273, NA,
                43636, Inf, NA),
              ncol = 3, byrow = TRUE)

#reclassify raster using reclassification matrix
landsat <- classify(landsat_masked, rcl = rcl_0)

landsat

#converting to reflectance based on known scale factors
#use scale factor for landsat 5
#multiply by 0.0000275
#add by -0.2
#this will convert current values to reflectance

landsat <- ((landsat*0.0000275) - 0.2)*100
```


## Classify image

#### Extract reflectance values for training data
We will load the shapefile identifying different locations within our study area as containing one of our 4 land cover types. We can then extract the spectral values at each site to create a data frame that relates land cover types to their spectral reflectance.

```{r include=TRUE}
#load up training data
training_data <- st_read("./data (1)/trainingdata.shp") %>%
  st_transform(., crs(landsat))

#pulling out the reflectance values for the pixels within each polygon
training_data_values <- extract(landsat, training_data, df = TRUE)

#creating data frame from sf object by dropping geom
training_data_attributes <- training_data %>%
  st_drop_geometry()

SB_training_data <- left_join(training_data_values, training_data_attributes,
                              by = c("ID" = "id")) %>%
  mutate(type = as.factor(type)) # converting land cover type to factor
```

#### Train decision tree classifier
To train our decision tree, we first need to establish our model formula (i.e. what our response and predictor variables are). The `rpart` function implements the [CART algorithm](https://medium.com/geekculture/decision-trees-with-cart-algorithm-7e179acee8ff). The `rpart` function needs to know the model formula and training data you would like to use. Because we are performing a classification, we set `method = "class"`. We also set `na.action = na.omit` to remove any pixels with `NA`s from the analysis.\

To understand how our decision tree will classify pixels, we can plot the results. The decision tree is comprised of a hierarchy of binary decisions. Each decision rule has 2 outcomes based on a conditional statement pertaining to values in each spectral band.   

```{r include=TRUE}
#establish model formula
SB_formula <- type ~ red + green + blue + NIR + SWIR1 + SWIR2

#training a decision tree classifier
SB_decision_tree <- rpart(formula = SB_formula,
                          data = SB_training_data,
                          method = "class",
                          na.action = na.omit)

prp(SB_decision_tree)
```

#### Apply decision tree
Now that we have created our decision tree, we can apply it to our entire image. The `terra` package includes a `predict()` function that allows us to apply a model to our data. In order for this to work properly, the names of the layers need to match the column names of the predictors we used to train our decision tree. The `predict()` function will return a raster layer with integer values. These integer values correspond to the *factor levels* in the training data. To figure out what category each integer corresponds to, we can inspect the levels of our training data. 

```{r include=TRUE}
SB_classification <- predict(landsat, SB_decision_tree, 
                             type = "class", na.rm = TRUE)

levels(SB_training_data$type)
```

#### Plot results
Now we can plot the results and check out our land cover map!
```{r}
tm_shape(SB_classification) +
  tm_raster(style = "cat",
            labels = c("green_vegetation",
                       "soil_dead_grass",
                       "urban",
                       "water"),
            title = "land cover")

```
